Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-02-01 Multiresolution_framework/ltx/manuscript.tex
edit to computational complexity

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}
 \end{equation}
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation in \citet{Dewar2009}. The complexity of each step of the algorithm is summarised in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method in \citet{Dewar2009} is also identified. 
\begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lllc}
\hline \hline
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&\eqref{eq:MRA-Xi0}, \eqref{eq:MRA-Xi1}&$O(Tn_x^2)$& -\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&\eqref{eq:epsilon0},\eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&\eqref{eq:A_theta}&$O(n_x^3)$&-\\
$\boldsymbol\Lambda_{\theta}$&\eqref{eq:Lambdatheta}&$O(n_x^2n_{\theta})$&-\\
$\hat{\boldsymbol\theta} $&\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&-\\
\hline \hline
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009}.} 
\label{table:MRA-ComputationalComplexity}
\end{center}
\end {table}
The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\boldsymbol\theta,\boldsymbol\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. The RTS  algorithm is given in Algorithm~\ref{alg:MRA-RTS} for completeness.

\renewcommand{\arraystretch}{1.7}
\begin{table*}[!ht]
\begin{tabular}{|c|}\hline
\multicolumn{1}{|p{16cm}|}{\textbf{1.} Forward initialization} \\ 
$\hat{\mathbf x}_0, \mathbf P_0$ \\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{2.} Forward iteration: for $t\in\left\lbrace 0,\cdots,T\right\rbrace $, calculate the predicted state and the predicted covariance matrix } \\
$\hat{\mathbf x}_{t+1}^{f-}=\mathbf A \hat{\mathbf x}_{t}^{f}$  \\ 
$\mathbf P_{t+1}^{f-}=\mathbf A \mathbf P_{t}^{f}\mathbf A^{\top}+\boldsymbol\Sigma_e$  \\
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix} \\
$\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2n_x} W_i^{(m)}\mathcal X_{i,t+1}^{f-} $ \\
$\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2n_x} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e$ \\ 
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman filter update equations} \\
$\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\varepsilon})^{-1}$\\
$\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})$\\
$\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-}$\\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{3.} Backward initialization}\\
 $\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T$\\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{4.} Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute the smoother gain, the smoothed state and the smoothed covariance matrix}\\  
$\mathcal S_{t}=\mathbf P_{t}^{f}\mathbf A^{\top}\left[ \mathbf P_{t +1}^{f-}\right]^{-1}$ \\
 $\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathcal S_t(\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{f-})$ \\
 $\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathcal S_t(\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{f-})\mathcal S_t^\top$\\  
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{5.} Compute the smoothed cross covariance matrix, initialization}\\
$\mathbf M_T=(\mathbf I-\mathcal K_T\mathbf C)\mathbf A\mathbf P_{T-1}^b$\\
\multicolumn{1}{|p{16cm}|}{For $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute}\\
$\mathbf M_t= \mathbf P_t^{f}\mathcal S_{t-1}^{\top}+\mathcal S_{t}(\mathbf M_{t+1}-\mathbf A\mathbf P_t^{f} )\mathcal S_{t-1}^{\top}$\\
\hline
\end{tabular}
\caption{\textbf{Algorithm for the RTS smoother} This table shows the steps in the  Rauch-Tung-Striebel smoother algorithm. The lower bound is maximised after each iteration to update the parameter estimates.}
\label{alg:MRA-RTS}
\end{table*}
\renewcommand{\arraystretch}{1}
%\cut{The second terms in  \eqref{eq:upsilon0}, \eqref{eq:upsilon1}, and \eqref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.}

\section{Results}\label{sec:MRA-results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using equations \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, allowing a comparison between true and estimated parameters. In doing so, the membrane time constant was set to $\tau = 10$~ms \citep{David2003}, and the sampling time was chosen ten times faster \citep{Stephan2008}, i.e. $T_s = 1$~ms, ensuring the discrepancy between the continuous neural field model and its discrete approximation is small.

>>>>

\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \citet{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}
 \end{equation}
\subsection{Computational complexity} 
All $n_x^2$ blocks of $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the implementation in \citet{Dewar2009}. The complexity of each step of the algorithm is summarised in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method in \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  
\begin {table}
\begin{center}
\scalebox{1}{\begin{tabular}{lllc}
\hline \hline
Variable &Equation&Order&Order obtained from \citet{Dewar2009}\\ \hline\\
$\Xi_0, \Xi_1$&\eqref{eq:MRA-Xi0}, \eqref{eq:MRA-Xi1}&$O(Tn_x^2)$& -\\
$\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$&\eqref{eq:epsilon0},\eqref{eq:epsilon1} &$O(n_x^2n_{\theta})$ &$O(n_x^2n_{\theta})+O(n_x^3)$ \\
$\boldsymbol\Upsilon$&\eqref{eq:Epsilon}&$O(n_x^2n_{\theta}^2)$&$O(n_x^4n_{\theta}^2)$\\
$\mathbf A(\boldsymbol\theta)$&\eqref{eq:A_theta}&$O(n_x^3)$&-\\
$\boldsymbol\Lambda_{\theta}$&\eqref{eq:Lambdatheta}&$O(n_x^2n_{\theta})$&-\\
$\hat{\boldsymbol\theta} $&\eqref{eq:MRA-thetahat}&$O(n_{\theta}^3)$&-\\
\hline \hline
\end{tabular}}
 \caption {{\bf The Computational complexity of the estimation algorithm}. A comparison between the algorithm proposed herein and that of \citet{Dewar2009}.} 
\label{table:MRA-ComputationalComplexity}
\end{center}
\end {table}

The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\boldsymbol\theta,\boldsymbol\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. The RTS  algorithm is given in Algorithm~\ref{alg:MRA-RTS} for completeness.

\renewcommand{\arraystretch}{1.7}
\begin{table*}[!ht]
\begin{tabular}{|c|}\hline
\multicolumn{1}{|p{16cm}|}{\textbf{1.} Forward initialization} \\ 
$\hat{\mathbf x}_0, \mathbf P_0$ \\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{2.} Forward iteration: for $t\in\left\lbrace 0,\cdots,T\right\rbrace $, calculate the predicted state and the predicted covariance matrix } \\
$\hat{\mathbf x}_{t+1}^{f-}=\mathbf A \hat{\mathbf x}_{t}^{f}$  \\ 
$\mathbf P_{t+1}^{f-}=\mathbf A \mathbf P_{t}^{f}\mathbf A^{\top}+\boldsymbol\Sigma_e$  \\
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix} \\
$\hat{\mathbf x}_{t+1}^{f-}=\sum_{i=0}^{2n_x} W_i^{(m)}\mathcal X_{i,t+1}^{f-} $ \\
$\mathbf P_{t +1}^{f-}=\sum_{i=0}^{2n_x} W_i^{(c)}(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})(\mathcal X_{i,t+1}^{f-}-\hat{\mathbf x}_{t +1}^{f-})^\top+\boldsymbol \Sigma_e$ \\ 
\multicolumn{1}{|p{16cm}|}{Compute the filter gain, the filtered state and the filtered covariance matrix using the standard Kalman filter update equations} \\
$\mathcal K_{t+1}=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\varepsilon})^{-1}$\\
$\hat{\mathbf x}_{t+1}^{f}=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})$\\
$\mathbf P_{t+1}^f=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-}$\\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{3.} Backward initialization}\\
 $\mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T$\\
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{4.} Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute the smoother gain, the smoothed state and the smoothed covariance matrix}\\  
$\mathcal S_{t}=\mathbf P_{t}^{f}\mathbf A^{\top}\left[ \mathbf P_{t +1}^{f-}\right]^{-1}$ \\
 $\hat{\mathbf x}_t^b=\hat{\mathbf x}_t^f+\mathcal S_t(\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{f-})$ \\
 $\mathbf P_{t}^{b}=\mathbf P_{t}^{f}+\mathcal S_t(\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{f-})\mathcal S_t^\top$\\  
\hline
\multicolumn{1}{|p{16cm}|}{\textbf{5.} Compute the smoothed cross covariance matrix, initialization}\\
$\mathbf M_T=(\mathbf I-\mathcal K_T\mathbf C)\mathbf A\mathbf P_{T-1}^b$\\
\multicolumn{1}{|p{16cm}|}{For $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute}\\
$\mathbf M_t= \mathbf P_t^{f}\mathcal S_{t-1}^{\top}+\mathcal S_{t}(\mathbf M_{t+1}-\mathbf A\mathbf P_t^{f} )\mathcal S_{t-1}^{\top}$\\
\hline
\end{tabular}
\caption{\textbf{Algorithm for the RTS smoother} This table shows the steps in the  Rauch-Tung-Striebel smoother algorithm. The lower bound is maximised after each iteration to update the parameter estimates.}
\label{alg:MRA-RTS}
\end{table*}
\renewcommand{\arraystretch}{1} 
\section{Results}\label{sec:MRA-results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using equations \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, allowing a comparison between true and estimated parameters. In doing so, the membrane time constant was set to $\tau = 10$~ms \citep{David2003}, and the sampling time was chosen ten times faster \citep{Stephan2008}, i.e. $T_s = 1$~ms, ensuring the discrepancy between the continuous neural field model and its discrete approximation is small.
